The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-06-01 source/main.tex

\begin{abstract}
Federated learning learns
a model jointly from a set of distributed workers without sharing their private
data. The characteristics of non-\emph{i.i.d.} data
across the network and low device participation bring significant challenges
in understanding the convergence of federated learning algorithms. Prior work
analyzed the convergence with restrained settings,
%either \emph{i.i.d.} data, or with full device participation, 
and often with suboptimal rates. 
In this work, we provide a
comprehensive convergence analysis on Federated Averaging (FedAvg),
a widely used framework in which SGD or Nesterov SGD updates are used for general convex problems with arbitrarily
heterogeneous data. 
We first show that the convergence rates of both algorithms enjoy a linear speedup w.r.t. the number of participating workers for all convex smooth problems. 
Furthermore, in the overparamterized setting, we show that the convergence of FedAvg is in fact exponential, also with a speedup factor linear in the number of workers. 
We propose a new accelerated federated learning algorithm that further improves this convergence rate with provable guarantees in the linear regression setting. 
Empirical studies of the algorithms in various settings have supported our theoretical results.
\end{abstract}

>>>>

\begin{abstract}
Federated learning learns
a model jointly from a set of distributed workers without sharing their private
data. The characteristics of non-\emph{i.i.d.} data
across the network and low device participation bring significant challenges
in understanding the convergence of federated learning algorithms. Prior work
analyzed the convergence with restrained settings,
%either \emph{i.i.d.} data, or with full device participation, 
and often with suboptimal rates. 
In this work, we provide a
comprehensive convergence analysis on Federated Averaging (FedAvg),
a widely used framework in which SGD or Nesterov SGD updates are used for general convex problems with arbitrarily
heterogeneous data. 
We first show that the convergence rates using both updates enjoy a linear speedup w.r.t. the number of participating workers for all convex smooth problems. 
In addition, in the overparamterized setting, we show that the convergence of FedAvg is in fact exponential, also with a speedup factor linear in the number of workers. 
Furthermore, we propose a new accelerated federated learning algorithm that further improves this convergence rate with provable guarantees in the linear regression setting. 
Empirical studies of the algorithms in various settings have supported our theoretical results.
\end{abstract}

### 2020-06-01 source/sec_conclude.tex

\section{Conclusions}
This paper provides a comprehensive analysis of the convergence rate of FedAvg
and its accelerated variants in a general federated learning problem with heterogeneous local data and partial participation. We show that both Nesterov accelerated FedAvg and FedAvg
can achieve {\small{$\cO(\frac{1}{\sqrt{NT}})$}} linear speedup convergence for convex smooth problems and {\small{$\cO(\frac{1}{NT})$}} convergence for strongly 
convex smooth problems. In addition, we show that the local steps for stronlgy convex and convex smooth problems can be as large as {\small{$\cO(\sqrt{\frac{T}{N}})$}}, which can save communication cost substantially comparing to prior results. 
Furthermore, this work also makes algorithmic contributions. We not only prove that FedAvg can achieve exponential convergence for overparameterized strongly convex smooth problems, but also propose the MaSS accelerated Fedavg algorithm, which has provable speedup in convergence rate over FedAvg on quadratic problems. Last but not least, we empirically
verify the linear speedup of FedAvg and Nesterov accelerated FedAvg for strongly convex, convex smooth, and linear regression problems. The empirical results are well-aligned with our theories.

>>>>

\section{Conclusions}
This paper provides a comprehensive analysis of the convergence rate of FedAvg
and its accelerated variants in a general federated learning problem with heterogeneous local data and partial participation. We show that both Nesterov accelerated FedAvg and FedAvg
can achieve {\small{$\cO(\frac{1}{\sqrt{NT}})$}} linear speedup convergence for convex smooth problems and {\small{$\cO(\frac{1}{NT})$}} convergence for strongly 
convex smooth problems. In addition, we show that the local steps for stronlgy convex and convex smooth problems can be as large as {\small{$\cO(\sqrt{\frac{T}{N}})$}}, which can substantially save communication cost comparing to prior results. 
Furthermore, this work also makes algorithmic contributions. We not only prove that FedAvg can achieve exponential convergence for overparameterized strongly convex smooth problems, but also propose the MaSS accelerated Fedavg algorithm, which has provable speedup in convergence rate over FedAvg on quadratic problems. Last but not least, we empirically
verify the linear speedup of FedAvg and Nesterov accelerated FedAvg for strongly convex, convex smooth, and linear regression problems. The empirical results are well-aligned with our theories.
